---
title: "Getting Started with fireR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fireR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>"
)
```

## Overview

**fireR** provides convenient access to USGS fire datasets and EPA/CEC
ecoregion boundaries:

- `get_mtbs()` / `read_mtbs()` for
  [MTBS (Monitoring Trends in Burn Severity)](https://www.mtbs.gov/) wildfire
  perimeter data
- `get_sefire()` for SE FireMap data products (Burn Severity mosaics
  2000–2022; Fire History, Burned Area Polygons, and Burned Area Rasters
  1994–2024)
- `get_nifc()` / `read_nifc()` for NIFC (National Interagency Fire Center)
  wildfire perimeters
- `get_fod()` / `read_fod()` for the USFS Fire Occurrence Database
  (FPA-FOD, 1992–2020)
- `get_wui()` for the USFS Wildland-Urban Interface (WUI) dataset
- `get_nal1eco()`, `get_nal2eco()`, `get_nal3eco()` for CEC North America
  ecoregion boundaries
- `get_usl3eco()`, `get_usl4eco()` for US EPA ecoregion boundaries

```{r setup}
library(fireR)
```

---

## A quick look

Most fireR functions pull large archives straight from the source, so the
examples below are not run when this site is built. The maps in this section,
however, are produced live from a single, modest download — the CEC North
America Level 1 ecoregion boundaries — to show what the returned spatial
objects look like with both base R `plot()` and **ggplot2**.

```{r gallery-load, message = FALSE, warning = FALSE, error = TRUE}
library(ggplot2)

# A single, reliably-hosted download (~tens of MB)
na_l1 <- get_nal1eco(verbose = FALSE)
```

Base R `plot()` of an `sf` object, coloured by ecoregion name:

```{r gallery-plot, error = TRUE, fig.width = 7, fig.height = 5, fig.alt = "Map of North America Level 1 ecoregions coloured by ecoregion name."}
plot(na_l1["NA_L1NAME"], main = "North America Level 1 Ecoregions", border = NA)
```

The same data rendered with **ggplot2** via `geom_sf()`:

```{r gallery-ggplot, error = TRUE, fig.width = 7, fig.height = 5, fig.alt = "ggplot2 map of North America Level 1 ecoregions filled by ecoregion name."}
ggplot(na_l1) +
  geom_sf(aes(fill = NA_L1NAME), color = NA) +
  guides(fill = "none") +
  labs(title = "North America Level 1 Ecoregions") +
  theme_minimal()
```

---

## MTBS fire perimeters

### All fires

With `output = "sf"`, `read_mtbs()` returns an **`sf`** object containing every
fire perimeter in the MTBS composite dataset — all years, all states.

```{r all-fires, eval = FALSE, purl = FALSE}
fires <- read_mtbs(output = "sf")
fires
#> Simple feature collection with 31,386 features and 22 fields
#> Geometry type: MULTIPOLYGON
#> Dimension:     XY
#> Bounding box:  xmin: -178.3 ymin: 17.9 xmax: -65.3 ymax: 71.4
#> CRS:           NAD83 / Conus Albers (EPSG:5070)
```

Plot the burn area acreage to get a quick sense of the data:

```{r plot-basic, eval = FALSE, purl = FALSE}
plot(fires["BurnBndAc"], main = "MTBS Fire Perimeters — All Years")
```

---

## Filtering by year

### Single year

Pass a single integer to keep only fires that started in that calendar year:

```{r single-year, eval = FALSE, purl = FALSE}
fires_2020 <- read_mtbs(years = 2020, output = "sf")
nrow(fires_2020)
#> [1] 246
```

### Year range

Use R's `:` operator to keep all fires within a contiguous span of years:

```{r year-range, eval = FALSE, purl = FALSE}
fires_recent <- read_mtbs(years = 2018:2023, output = "sf")
nrow(fires_recent)
#> [1] 1328
```

Quickly visualise where the major fires of the last decade fell:

```{r map-recent, eval = FALSE, purl = FALSE}
library(sf)

plot(st_geometry(fires_recent), col = "#E25822AA", border = NA,
     main = "MTBS Fire Perimeters 2018–2023")
```

### Specific years only

Supply a vector of individual years to return only fires from those exact years
(non-contiguous):

```{r specific-years, eval = FALSE, purl = FALSE}
fires_sel <- read_mtbs(years = c(2000, 2010, 2020), output = "sf")
```

---

## Output formats

### terra::SpatVector

Set `output = "vect"` (or `"terra"`) to receive a `terra::SpatVector` instead
of an `sf` object — handy when the rest of your workflow uses **terra**:

```{r terra-output, eval = FALSE, purl = FALSE}
library(terra)

fires_vect <- read_mtbs(years = 2020:2023, output = "vect")
class(fires_vect)
#> [1] "SpatVector"
#> attr(,"package")
#> [1] "terra"

plot(fires_vect, "BurnBndAc", main = "Recent Fire Perimeters (terra)")
```

### Attribute table only

Set `geometry = FALSE` to drop geometry and get a plain `data.frame`.
Useful when you only need the metadata (fire name, year, acreage, etc.):

```{r no-geometry, eval = FALSE, purl = FALSE}
tbl <- read_mtbs(geometry = FALSE)
head(tbl[, c("Incid_Name", "Ig_Date", "BurnBndAc", "Incid_Type")])
#>     Incid_Name   Ig_Date BurnBndAc Incid_Type
#> 1  LOWDEN FIRE 1984-07-01   4537.35   Wildfire
#> 2  EUREKA FIRE 1984-08-15  11202.06   Wildfire
#> ...
```

---

## Caching downloads

The MTBS ZIP archive is ~360 MB. Download it once with `get_mtbs()`, then read
from disk with `read_mtbs()` on every subsequent call.

### Default cache directory

Pass `cache = TRUE` to `read_mtbs()` to look for the ZIP in
`tools::R_user_dir("fireR", "cache")` — a platform-appropriate user directory
that persists between R sessions:

```{r cache-default, eval = FALSE, purl = FALSE}
# Download once to the user cache directory
get_mtbs(directory = tools::R_user_dir("fireR", "cache"))

# Read from cache on every subsequent call
fires <- read_mtbs(cache = TRUE)
```

### Custom cache directory

Supply a directory path to both functions to control where the file is stored:

```{r cache-custom, eval = FALSE, purl = FALSE}
get_mtbs(directory = "~/data/mtbs_cache")
fires <- read_mtbs(cache = "~/data/mtbs_cache")
```

### Force a fresh download

If the USGS releases an updated dataset, use `overwrite = TRUE` on `get_mtbs()`
to bypass the cache and re-download:

```{r cache-overwrite, eval = FALSE, purl = FALSE}
get_mtbs(directory = tools::R_user_dir("fireR", "cache"), overwrite = TRUE)
fires <- read_mtbs(cache = TRUE)
```

---

## Quiet mode

Progress messages are printed by default. Suppress them with `verbose = FALSE`:

```{r quiet, eval = FALSE, purl = FALSE}
fires <- read_mtbs(years = 2022, verbose = FALSE)
```

---

## SE FireMap

`get_sefire()` downloads four SE FireMap data products via the `dataset`
argument: annual Burn Severity mosaics (year-based, the default) and three
single-file products covering 1994–2024 — Fire History, Burned Area Polygons,
and Burned Area Rasters.

### Burn Severity mosaics

`dataset = "Burn Severity"` (the default) downloads one Annual Burn Severity
Mosaic ZIP per year for the southeastern United States (2000–2022).  Pass a
single year, a range, or a vector of specific years:

```{r sefire-single, eval = FALSE, purl = FALSE}
# Single year
zip_path <- get_sefire(years = 2020)
```

```{r sefire-range, eval = FALSE, purl = FALSE}
# Contiguous range
zip_paths <- get_sefire(years = 2015:2020, directory = "data/sefire")
```

```{r sefire-specific, eval = FALSE, purl = FALSE}
# Specific years only
zip_paths <- get_sefire(years = c(2000, 2010, 2020))
```

### Single-file datasets (1994–2024)

The remaining three datasets are each a single geodatabase ZIP covering
1994–2024.  The `years` argument does not apply to them:

```{r sefire-fire-history, eval = FALSE, purl = FALSE}
# Fire History
zip_path <- get_sefire(dataset = "Fire History", directory = "data/sefire")
```

```{r sefire-burned-area-polys, eval = FALSE, purl = FALSE}
# Burned Area Polygons
zip_path <- get_sefire(dataset = "Burned Area Polygons", directory = "data/sefire")
```

```{r sefire-burned-area-rasters, eval = FALSE, purl = FALSE}
# Burned Area Rasters
zip_path <- get_sefire(dataset = "Burned Area Rasters", directory = "data/sefire")
```

---

## NIFC wildfire perimeters

`get_nifc()` downloads the NIFC wildfire perimeters ZIP; `read_nifc()` reads
and filters it, with optional year filtering via the `FireYear` column:

```{r nifc, eval = FALSE, purl = FALSE}
# Download once
get_nifc(directory = "data/nifc")

# Read all perimeters, or filter by year
perims      <- read_nifc(cache = "data/nifc", output = "sf")
perims_2021 <- read_nifc(cache = "data/nifc", years = 2021)
```

---

## USFS Fire Occurrence Database (FPA-FOD)

`get_fod()` downloads the FPA-FOD GeoPackage ZIP; `read_fod()` reads and
filters it. The dataset covers 1992–2020 (`FIRE_YEAR` column):

```{r fod, eval = FALSE, purl = FALSE}
get_fod(directory = "data/fod")

fod_recent <- read_fod(cache = "data/fod", years = 2015:2020)
```

---

## Wildland-Urban Interface (WUI)

`get_wui()` downloads the USFS Wildland-Urban Interface dataset, which
delineates areas where structures meet or intermingle with undeveloped
wildland vegetation. The ZIP is **very large (~4.65 GB)**, and the function
warns about this before downloading. The recommended pattern is to run the
download in a background R session with `callr::r_bg()` so your console stays
free:

```{r wui, eval = FALSE, purl = FALSE}
# Recommended: run in a background session (4.65 GB download)
bg <- callr::r_bg(function() fireR::get_wui(directory = "data/wui"))
bg$wait()

# Or download directly in the current session
zip_path <- get_wui(directory = "data/wui")
```

---

## Ecoregion boundaries

### North America (CEC Levels 1–3)

The Commission for Environmental Cooperation (CEC) ecoregion framework
divides North America into hierarchical ecological units.

```{r na-eco, eval = FALSE, purl = FALSE}
# Level 1 — broadest continental divisions
na_l1 <- get_nal1eco()

# Level 2 — finer continental subdivisions
na_l2 <- get_nal2eco()

# Level 3 — finest continental scale (= US EPA Level III)
na_l3 <- get_nal3eco()

plot(na_l1["NA_L1NAME"], main = "North America Level 1 Ecoregions")
```

### US EPA (Levels 3–4)

US EPA ecoregions are available at Levels 3 and 4, with an option to include
state boundaries in the polygons.

```{r us-eco, eval = FALSE, purl = FALSE}
# Level 3 — without state boundaries (default)
us_l3 <- get_usl3eco()

# Level 3 — with state boundaries dissolved in
us_l3_states <- get_usl3eco(state = TRUE)

# Level 4 — finest US subdivisions
us_l4 <- get_usl4eco()
us_l4_states <- get_usl4eco(state = TRUE)
```

All ecoregion functions accept `output = "vect"` for a `terra::SpatVector`
and a `cache` argument to persist downloads across sessions:

```{r eco-cache, eval = FALSE, purl = FALSE}
us_l3 <- get_usl3eco(output = "vect", cache = TRUE)
```

---

## Working with the data

Once you have an `sf` object, the full **sf** and **dplyr** ecosystem is
available to you.

### Key MTBS columns

| Column | Description |
|---|---|
| `Incid_Name` | Name of the fire event |
| `Ig_Date` | Ignition date (YYYY-MM-DD) |
| `BurnBndAc` | Burned area in acres |
| `Incid_Type` | Incident type (Wildfire, Prescribed Fire, etc.) |
| `irwinID` | Unique IRWIN identifier |
| `geometry` | Fire perimeter polygon(s) |

### Example: largest fires since 2000

```{r largest-fires, eval = FALSE, purl = FALSE}
library(dplyr)

fires_2000s <- read_mtbs(years = 2000:2023, output = "sf")

top10 <- fires_2000s |>
  slice_max(BurnBndAc, n = 10) |>
  select(Incid_Name, Ig_Date, BurnBndAc) |>
  st_drop_geometry()

top10
```

### Example: area burned and fires per year

```{r fires-load, eval = FALSE, purl = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(sf)

fires_all <- read_mtbs(output = "sf")
```

```{r plot-area-burned, eval = FALSE, purl = FALSE}
fires_all |>
  st_drop_geometry() |>
  mutate(year = as.integer(substr(Ig_Date, 1, 4))) |>
  group_by(year) |>
  summarize(area_burned = sum(BurnBndAc)) |>
  ggplot(aes(year, area_burned)) +
  geom_col(fill = "#8B1A1A") +
  labs(title = "Area Burned by Year",
       x = "Year",
       y = "Area Burned (Acres)") +
  theme_classic()
```

```{r plot-fires-per-year, eval = FALSE, purl = FALSE}
fires_all |>
  st_drop_geometry() |>
  mutate(year = as.integer(substr(Ig_Date, 1, 4))) |>
  count(year) |>
  ggplot(aes(year, n)) +
  geom_col(fill = "#8B1A1A") +
  labs(title = "Number of Fires by Year",
       x = "Year",
       y = "Number of Fires") +
  theme_classic()
```

---

## About MTBS

The **Monitoring Trends in Burn Severity** programme is a joint USGS / USFS
initiative that maps the location, extent, and burn severity of all large
wildfires (>1 000 acres in the western US; >500 acres in the eastern US) across
the conterminous USA, Alaska, Hawaii, and Puerto Rico from **1984 to the
present**.

More information: <https://www.mtbs.gov/>
