---
title: "Getting Started with fireR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fireR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

**fireR** provides convenient access to USGS fire datasets and EPA/CEC
ecoregion boundaries:

- `get_mtbs()` / `read_mtbs()` for
  [MTBS (Monitoring Trends in Burn Severity)](https://www.mtbs.gov/) wildfire
  perimeter data
- `get_sefire()` for SE FireMap Annual Burn Severity Mosaics
- `get_nal1eco()`, `get_nal2eco()`, `get_nal3eco()` for CEC North America
  ecoregion boundaries
- `get_usl3eco()`, `get_usl4eco()` for US EPA ecoregion boundaries

```{r setup}
library(fireR)
```

---

## MTBS fire perimeters

### All fires

With `output = "sf"`, `read_mtbs()` returns an **`sf`** object containing every
fire perimeter in the MTBS composite dataset — all years, all states.

```{r all-fires}
fires <- read_mtbs(output = "sf")
fires
#> Simple feature collection with 31,386 features and 22 fields
#> Geometry type: MULTIPOLYGON
#> Dimension:     XY
#> Bounding box:  xmin: -178.3 ymin: 17.9 xmax: -65.3 ymax: 71.4
#> CRS:           NAD83 / Conus Albers (EPSG:5070)
```

Plot the burn area acreage to get a quick sense of the data:

```{r plot-basic}
plot(fires["BurnBndAc"], main = "MTBS Fire Perimeters — All Years")
```

---

## Filtering by year

### Single year

Pass a single integer to keep only fires that started in that calendar year:

```{r single-year}
fires_2020 <- read_mtbs(years = 2020, output = "sf")
nrow(fires_2020)
#> [1] 246
```

### Year range

Use R's `:` operator to keep all fires within a contiguous span of years:

```{r year-range}
fires_recent <- read_mtbs(years = 2018:2023, output = "sf")
nrow(fires_recent)
#> [1] 1328
```

Quickly visualise where the major fires of the last decade fell:

```{r map-recent}
library(sf)

plot(st_geometry(fires_recent), col = "#E25822AA", border = NA,
     main = "MTBS Fire Perimeters 2018–2023")
```

### Specific years only

Supply a vector of individual years to return only fires from those exact years
(non-contiguous):

```{r specific-years}
fires_sel <- read_mtbs(years = c(2000, 2010, 2020), output = "sf")
```

---

## Output formats

### terra::SpatVector

Set `output = "vect"` (or `"terra"`) to receive a `terra::SpatVector` instead
of an `sf` object — handy when the rest of your workflow uses **terra**:

```{r terra-output}
library(terra)

fires_vect <- read_mtbs(years = 2020:2023, output = "vect")
class(fires_vect)
#> [1] "SpatVector"
#> attr(,"package")
#> [1] "terra"

plot(fires_vect, "BurnBndAc", main = "Recent Fire Perimeters (terra)")
```

### Attribute table only

Set `geometry = FALSE` to drop geometry and get a plain `data.frame`.
Useful when you only need the metadata (fire name, year, acreage, etc.):

```{r no-geometry}
tbl <- read_mtbs(geometry = FALSE)
head(tbl[, c("Incid_Name", "Ig_Date", "BurnBndAc", "Incid_Type")])
#>     Incid_Name   Ig_Date BurnBndAc Incid_Type
#> 1  LOWDEN FIRE 1984-07-01   4537.35   Wildfire
#> 2  EUREKA FIRE 1984-08-15  11202.06   Wildfire
#> ...
```

---

## Caching downloads

The MTBS ZIP archive is ~100 MB. Download it once with `get_mtbs()`, then read
from disk with `read_mtbs()` on every subsequent call.

### Default cache directory

Pass `cache = TRUE` to `read_mtbs()` to look for the ZIP in
`tools::R_user_dir("fireR", "cache")` — a platform-appropriate user directory
that persists between R sessions:

```{r cache-default}
# Download once to the user cache directory
get_mtbs(directory = tools::R_user_dir("fireR", "cache"))

# Read from cache on every subsequent call
fires <- read_mtbs(cache = TRUE)
```

### Custom cache directory

Supply a directory path to both functions to control where the file is stored:

```{r cache-custom}
get_mtbs(directory = "~/data/mtbs_cache")
fires <- read_mtbs(cache = "~/data/mtbs_cache")
```

### Force a fresh download

If the USGS releases an updated dataset, use `overwrite = TRUE` on `get_mtbs()`
to bypass the cache and re-download:

```{r cache-overwrite}
get_mtbs(directory = tools::R_user_dir("fireR", "cache"), overwrite = TRUE)
fires <- read_mtbs(cache = TRUE)
```

---

## Quiet mode

Progress messages are printed by default. Suppress them with `verbose = FALSE`:

```{r quiet}
fires <- read_mtbs(years = 2022, verbose = FALSE)
```

---

## SE FireMap mosaics

`get_sefire()` downloads Annual Burn Severity Mosaic ZIP archives for the
southeastern United States (2000–2022).  Pass a single year, a range, or a
vector of specific years:

```{r sefire-single}
# Single year
zip_path <- get_sefire(2020)
```

```{r sefire-range}
# Contiguous range
zip_paths <- get_sefire(2015:2020, directory = "data/sefire")
```

```{r sefire-specific}
# Specific years only
zip_paths <- get_sefire(c(2000, 2010, 2020))
```

---

## Ecoregion boundaries

### North America (CEC Levels 1–3)

The Commission for Environmental Cooperation (CEC) ecoregion framework
divides North America into hierarchical ecological units.

```{r na-eco}
# Level 1 — broadest continental divisions
na_l1 <- get_nal1eco()

# Level 2 — finer continental subdivisions
na_l2 <- get_nal2eco()

# Level 3 — finest continental scale (= US EPA Level III)
na_l3 <- get_nal3eco()

plot(na_l1["NA_L1NAME"], main = "North America Level 1 Ecoregions")
```

### US EPA (Levels 3–4)

US EPA ecoregions are available at Levels 3 and 4, with an option to include
state boundaries in the polygons.

```{r us-eco}
# Level 3 — without state boundaries (default)
us_l3 <- get_usl3eco()

# Level 3 — with state boundaries dissolved in
us_l3_states <- get_usl3eco(state = TRUE)

# Level 4 — finest US subdivisions
us_l4 <- get_usl4eco()
us_l4_states <- get_usl4eco(state = TRUE)
```

All ecoregion functions accept `output = "vect"` for a `terra::SpatVector`
and a `cache` argument to persist downloads across sessions:

```{r eco-cache}
us_l3 <- get_usl3eco(output = "vect", cache = TRUE)
```

---

## Working with the data

Once you have an `sf` object, the full **sf** and **dplyr** ecosystem is
available to you.

### Key MTBS columns

| Column | Description |
|---|---|
| `Incid_Name` | Name of the fire event |
| `Ig_Date` | Ignition date (YYYY-MM-DD) |
| `BurnBndAc` | Burned area in acres |
| `Incid_Type` | Incident type (Wildfire, Prescribed Fire, etc.) |
| `irwinID` | Unique IRWIN identifier |
| `geometry` | Fire perimeter polygon(s) |

### Example: largest fires since 2000

```{r largest-fires}
library(dplyr)

fires_2000s <- read_mtbs(years = 2000:2023, output = "sf")

top10 <- fires_2000s |>
  slice_max(BurnBndAc, n = 10) |>
  select(Incid_Name, Ig_Date, BurnBndAc) |>
  st_drop_geometry()

top10
```

### Example: area burned and fires per year

```{r fires-load, eval = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(sf)

fires_all <- read_mtbs(output = "sf")
```

```{r plot-area-burned, eval = FALSE}
fires_all |>
  st_drop_geometry() |>
  mutate(year = as.integer(substr(Ig_Date, 1, 4))) |>
  group_by(year) |>
  summarize(area_burned = sum(BurnBndAc)) |>
  ggplot(aes(year, area_burned)) +
  geom_col(fill = "#8B1A1A") +
  labs(title = "Area Burned by Year",
       x = "Year",
       y = "Area Burned (Acres)") +
  theme_classic()
```

```{r plot-fires-per-year, eval = FALSE}
fires_all |>
  st_drop_geometry() |>
  mutate(year = as.integer(substr(Ig_Date, 1, 4))) |>
  count(year) |>
  ggplot(aes(year, n)) +
  geom_col(fill = "#8B1A1A") +
  labs(title = "Number of Fires by Year",
       x = "Year",
       y = "Number of Fires") +
  theme_classic()
```

---

## About MTBS

The **Monitoring Trends in Burn Severity** programme is a joint USGS / USFS
initiative that maps the location, extent, and burn severity of all large
wildfires (>1 000 acres in the western US; >500 acres in the eastern US) across
the conterminous USA, Alaska, Hawaii, and Puerto Rico from **1984 to the
present**.

More information: <https://www.mtbs.gov/>
